LayeredTypes

### 2022-10-25 paper/paper.tex
Slight adaptions to grammar
First brief architecture explanation

@@

\begin{figure}[ht!]
\begin{align*}
	id \in \texttt{Ident}\ 	&::=\ [a...zA...Z]+ \\
	v \in \texttt{Var}\ 	&::=\ id \\
	f \in \texttt{Fun}\ 	&::=\ id_1\ id_2\ \{\ s\ \}\\
	e \in \texttt{Expr}\ 	&::=\ v\ |\ c\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ f(e) \\
	c \in \texttt{Const}\ 	&::=\ \texttt{True}\ |\ \texttt{False}\ |\ \mathbb{Z} \\
	s \in \texttt{Stmt}\ 	&::=\ v\ :=\ e\ |\ \texttt{if}\ v\ \texttt{then}\ s\ \texttt{else}\ s\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ s\ |\ l \\
	l \in \texttt{Layer}\ 	&::=\ id_1::id_2::\texttt{String} \\
	\circledplus\ 		&::=\ \texttt{+}\ |\ \texttt{-} \\
	\circleddot\ 		&::=\ \texttt{!=}\ |\ \texttt{==}\ |\ \texttt{<} \ |\ \texttt{>} \ |\ \texttt{<=} \ |\ \texttt{>=} \\
\end{align*}
\caption{Syntax of the simple functional language}
\label{fig:lang}
\end{figure}

@@

\begin{enumerate}
	\item A parser that parses the input program and creates an abstract syntax tree (AST) of the program.
	\item A layered type checker that checks the defined type layers for the tree incrementally.
	\item A code generator that generates the corresponding code for the input program.
\end{enumerate}

>>>>

% Shorthand command for Lark
\newcommand{\Lark}[0]{\textsc{Lark}}

@@

\begin{figure}[ht!]
\begin{align*}
	id \in \texttt{Ident}\ 	&::=\ [a...zA...Z]+ \\
	v \in \texttt{Var}\ 	&::=\ id \\
	f \in \texttt{Fun}\ 	&::=\ id_1\ id_2\ \{\ s\ \}\\
	e \in \texttt{Expr}\ 	&::=\ v\ |\ c\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ f(e) \\
	\texttt{CustomExpr}\ 	&::=\ \texttt{String}\\
	c \in \texttt{Const}\ 	&::=\ \texttt{True}\ |\ \texttt{False}\ |\ \mathbb{Z} \\
	\texttt{Assign}\ 	&::=\ v\ :=\ e\ |\ v\ :=\ \texttt{CustomExpr}\ \\
	s \in \texttt{Stmt}\ 	&::=\ \texttt{Assign}\ |\ \texttt{if}\ v\ \texttt{then}\ s\ \texttt{else}\ s\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ s\ |\ \ell\ |\ s_1\ s_2\ \\
	\ell \in \texttt{Layer}\ 	&::=\ id_1::id_2::\texttt{String} \\
	\circledplus\ 		&::=\ \texttt{+}\ |\ \texttt{-} \\
	\circleddot\ 		&::=\ \texttt{!=}\ |\ \texttt{==}\ |\ \texttt{<} \ |\ \texttt{>} \ |\ \texttt{<=} \ |\ \texttt{>=} \\
\end{align*}
\caption{Syntax of the simple functional language}
\label{fig:lang}
\end{figure}

While our language allows to define functions on its own, we will further allow the usage of function names that may have not been defined in the language itself. To allow this, the user can create a separate Python file that contains the definitions of additional functions. Our interpreter will then forward the calls to these functions to the Python interpreter. This allows us to easily access more complicated functionality such as the \texttt{print} function.

@@

\begin{enumerate}
	\item A parser that parses the input program and creates an abstract syntax tree (AST) of the program.
	\item A layered type checker that checks the defined type layers for the tree incrementally.
	\item An interpreter that executes the program.
\end{enumerate}

To parse a input program, we will use the Python library \Lark{} \cite{Lark}. This library allows us to define a grammar for the language and then parse a program according to this grammar. All further processing will be performed on the resulting AST. To process and execute the AST we will use the Visitors and Transformers provided by the \Lark{} library.

After initially parsing the AST, it will contain explicit nodes for each of the defined layers. As a first processing step we use a Transformer to collect these layers and remove the corresponding nodes from the AST. The implementation of each node has to be provided by the user as a separate Python module. Our language Interpreter will then dynamically load the corresponding module and use it to check the layer. To interact with our layered type checker correctly the module has to implement the following functions:

\begin{itemize}
	\item \texttt{depends\_on} -- This function returns a list of all layers that this layer depends on. This is used to ensure that the layers are checked in the correct order.
	\item \texttt{parse\_type} -- This function may be used to define and parse custom data types that are defined usinf the $CustomExpr$ rule in the grammar. \todo{Implementation details -- When and how is this function called?}
	\item \texttt{check\_layer} -- This function is used to check the layer. It is passed the AST of the program with all the annotated information from the previous layers. It should return a boolean value that indicates whether the layer is valid or not.
\end{itemize}

Using the \texttt{depends\_on} function, the layered type checker will first build a dependency graph of all layers. It will then topologically sort the layers and check them in the correct order. Once all layers have been checked successfully, the interpreter will execute the program.
